A comprehensive framework to capture the arcana of neuroimaging analysis

### 2018-10-12 manuscript/manuscript.tex
more automated replacing of citations

Despite the availability of well-established neuroimaging analysis
packages (Cox 1996; \cite{smith_advances_2004}; \cite{friston_statistical_2007}; Tournier
et. al. 2012), the arcana of neuroimaging analysis is substantial due to
the range of available tools, tuneable parameters, and imaging sequences
involved (\cite{cusack_automatic_2015}). The distribution of complete
``end-to-end'' workflows, from acquired data to publication results, is
necessary for routine reproduction because of the effort required to
accurately exactly reimplement such analyses (Kennedy 2018). It is also
difficult to adapt existing workflows to new studies without detailed
knowledge of their design. Therefore, flexible, portable and complete
workflows are important to promote reproduction and code reuse in
neuroimaging research.

@@

While neuroimaging analyses are generally amenable to standardisation
(Kennedy 2018), minor modifications are often required to accommodate
idiosyncrasies of the acquisition protocols in use at different sites
(Esteban et al. 2018). Therefore, workflows may require conditional
logic in construction or execution to be portable. Nipype is a flexible
Python framework for neuroimaging analysis in which workflows are
constructed programmatically in Python (Gorgolewski et al. 2011).
Programmatic construction allows for rich control-flow logic that is not
readily available in alternative workflow frameworks (Cusack et. al.
2015; \cite{achterberg_fastr:_2016}; \cite{amstutz_common_2016}), and has been used
to implement workflows that are robust to differences in fMRI protocols
across a large number of sites (Esteban et al. 2018).

The trend towards large multi-site and multi-contrast datasets collected
over a number of years (Van Essen et. al. 2012; \cite{thompson_enigma_2014};
\cite{sudlow_uk_2015}) presents additional challenges to workflow design.
Analysis packages are constantly being developed and improved, so the
state-of-the-art workflow for a particular analysis can change over
time. Therefore, it is challenging to ensure workflows are applied
consistently over the course of long studies (\cite{cusack_automatic_2015}).

@@

The utility of the Arcana framework is demonstrated by the
implementation of analysis suites for T1, T2* and diffusion weighted MRI
(dMRI) data and the application of dMRI tractogram (Tournier 2012) and
composite vein image (Ward 2018) workflows to data collected from a
healthy subject.

@@

Software versions are managed by the processor if environment modules
(Furlani 1991) are installed on the system. These are loaded and
unloaded before the interface of a workflow node is run based on the
requirements specified for the node during construction of the pipeline
(see \emph{Pipeline constructors}). Maps from requirement names to
module names in the processing environment can be passed to the
Processor at initialisation. For plugins that submit jobs to HPC
schedulers (e.g. SLURM, SGE), node memory and wall-time requirements are
inserted into jobscripts where required.

@@

The \emph{T2starwT1wStudy} class implements an algorithm to generate
\emph{composite vein images} from the combination of QSM and SWI
contrasts derived from the T2*-weighted acquisition with an vein atlas
registered to the T1-weighted acquisition (Ward 2018) (Figure 10).

@@

To test the analysis workflows implemented in ArcanaPaper, a healthy
volunteer was scanned using a 3T Siemens Skyra with a 32-channel head
and neck coil. The protocol was a T1-weighted MPRAGE (1mm contiguous,
matrix size 256x240x192, FOV 256x240x192, TE = 2.13ms, TR, = 2300ms, TI
= 900ms, bandwidth=230Hz/pixel), GRE (1.8 mm contiguous, matrix size
256x232x72, FOV 230x208x130, TE = 20ms, TR, = 30ms,
bandwidth=120Hz/pixel), diffusion MRI (1.2 mm contiguous, matrix size
110x100x60, FOV 256x240x192, TE = 95ms, TR, = 8200ms, 33 diffusion
directions with b=1500 mm2/s and 3 b=0, bandwidth=781Hz/pixel)
(\textbf{cite Lifespan paper ??}). Methods for Figure 11, 12, and 13 in
the ArcanaPaper class were applied to the acquired dataset to using the
\emph{SlurmProcessor} to submit jobs to the the MASSIVE HPC cluster
(Goscinski 2014) in order to generate the figures.

Figure \ref{fig:veins} displays composite vein images and vein masks for the healthy
volunteer along with the SWI and QSM intermediate derivatives and the
input vein atlas. The derived vein images are comparable to those
generated by the original implementation (Ward 2018). The complete
workflow took ?? min to complete over ?? compute cores.

@@

Software frameworks (Yacoub and Ammar 2004) have been successful in
improving code quality and efficiency of development in a variety of
contexts (\cite{moore_professional_2008}; White 2012; \cite{abadi_tensorflow:_2016}). By
factoring out common elements, only features that are specific to the
given application need to be implemented by the analysis designer, and
the common elements become battle hardened through repeated use. Arcana
handles many of the menial tasks involved with workflow implementation,
such as data retrieval and storage, format conversions, and provenance,
reducing the time and effort required to implement robust workflows from
acquired data to publication results.

An oft-repeated mantra in the open-source software movement dubbed
Linus' Law is that "given a large enough beta-tester and co-developer
base, almost every problem will be characterized quickly and the fix
obvious to someone" or more compactly, "given enough eyeballs, all
\href{https://en.wikipedia.org/wiki/Software_bug}{bugs} are shallow"
(Raymond 1999). Given the size of the neuroimaging research community,
there are a large number of potential beta-testers and co-developers.
However, it has been difficult for researchers to collaborate on the
same code base due to slight differences in acquisition protocols,
storage conventions, researcher preferences, and study requirements.

@@

The abstraction of data and repositories in Arcana enables the same
workflow implementation to be applied to datasets stored in BIDS format
or XNAT repositories. A single code-base can therefore be containerized
into BIDS apps or XNAT pipelines without adaptation, helping to form a
bridge between the two communities of users and developers. Alternative
data storage systems (Book et. al. 2009; Das et. al. 2011, Scott et. al
2011), can be integrated into Arcana by overriding a small number of
methods from the Repository abstract base class. Repository modules
could also be created for data portals such as \emph{DataLad} (Halchenko
2018) in order to take advantage of the range of platforms they support.
Implementing analyses in Arcana therefore enables researchers and
research groups to easily migrate their workflows between storage
platforms, and not risk being locked in to a particular technology.

@@

Ensuring that the consistent versions of external tools are used
throughout the analysis is important to avoid introducing biases due to
algorithm updates. In systems with environment modules (Furlani 1991)
installed, Arcana loads and unloads the required modules before and
after each node is executed. When running Arcana within a container,
environment modules and software versions can be installed inside the
container giving exact control over the versions used. To these ends, a
Docker container is available on Docker Hub,
(\href{http://hub.docker.com/r/monashbi/banana}{{http://hub.docker.com/r/monashbi/banana}}),
which can be used as a base for neuroimaging analysis containers.

>>>>

Despite the availability of well-established neuroimaging analysis
packages (\cite{cox_afni:_1996}; \cite{smith_advances_2004}; \cite{friston_statistical_2007};
Tournier et. al. 2012), the arcana of neuroimaging analysis is substantial due to
the range of available tools, tuneable parameters, and imaging sequences
involved (\cite{cusack_automatic_2015}). The distribution of complete
``end-to-end'' workflows, from acquired data to publication results, is
necessary for routine reproduction because of the effort required to
accurately exactly reimplement such analyses (\cite{kennedy_neuroimaging_2018}). It is also
difficult to adapt existing workflows to new studies without detailed
knowledge of their design. Therefore, flexible, portable and complete
workflows are important to promote reproduction and code reuse in
neuroimaging research.

@@

While neuroimaging analyses are generally amenable to standardisation
(\cite{kennedy_neuroimaging_2018}), minor modifications are often required to accommodate
idiosyncrasies of the acquisition protocols in use at different sites
(Esteban et al. 2018). Therefore, workflows may require conditional
logic in construction or execution to be portable. Nipype is a flexible
Python framework for neuroimaging analysis in which workflows are
constructed programmatically in Python (Gorgolewski et al. 2011).
Programmatic construction allows for rich control-flow logic that is not
readily available in alternative workflow frameworks (Cusack et. al. 2015; \cite{achterberg_fastr:_2016}; \cite{amstutz_common_2016}), and has been used
to implement workflows that are robust to differences in fMRI protocols
across a large number of sites (Esteban et al. 2018).

The trend towards large multi-site and multi-contrast datasets collected
over a number of years (Van et. al. 2012; \cite{thompson_enigma_2014};
\cite{sudlow_uk_2015}) presents additional challenges to workflow design.
Analysis packages are constantly being developed and improved, so the
state-of-the-art workflow for a particular analysis can change over
time. Therefore, it is challenging to ensure workflows are applied
consistently over the course of long studies (\cite{cusack_automatic_2015}).

@@

The utility of the Arcana framework is demonstrated by the
implementation of analysis suites for T1, T2* and diffusion weighted MRI
(dMRI) data and the application of dMRI tractogram (\cite{tournier_mrtrix:_2012}) and
composite vein image (\cite{ward_combining_2018}) workflows to data collected from a
healthy subject.

@@

Software versions are managed by the processor if environment modules
(\cite{furlani_modules:_1991}) are installed on the system. These are loaded and
unloaded before the interface of a workflow node is run based on the
requirements specified for the node during construction of the pipeline
(see \emph{Pipeline constructors}). Maps from requirement names to
module names in the processing environment can be passed to the
Processor at initialisation. For plugins that submit jobs to HPC
schedulers (e.g. SLURM, SGE), node memory and wall-time requirements are
inserted into jobscripts where required.

@@

The \emph{T2starwT1wStudy} class implements an algorithm to generate
\emph{composite vein images} from the combination of QSM and SWI
contrasts derived from the T2*-weighted acquisition with an vein atlas
registered to the T1-weighted acquisition (\cite{ward_combining_2018}) (Figure 10).

@@

To test the analysis workflows implemented in ArcanaPaper, a healthy
volunteer was scanned using a 3T Siemens Skyra with a 32-channel head
and neck coil. The protocol was a T1-weighted MPRAGE (1mm contiguous,
matrix size 256x240x192, FOV 256x240x192, TE = 2.13ms, TR, = 2300ms, TI
= 900ms, bandwidth=230Hz/pixel), GRE (1.8 mm contiguous, matrix size
256x232x72, FOV 230x208x130, TE = 20ms, TR, = 30ms,
bandwidth=120Hz/pixel), diffusion MRI (1.2 mm contiguous, matrix size
110x100x60, FOV 256x240x192, TE = 95ms, TR, = 8200ms, 33 diffusion
directions with b=1500 mm2/s and 3 b=0, bandwidth=781Hz/pixel)
(\textbf{cite Lifespan paper ??}). Methods for Figure 11, 12, and 13 in
the ArcanaPaper class were applied to the acquired dataset to using the
\emph{SlurmProcessor} to submit jobs to the the MASSIVE HPC cluster
(\cite{goscinski_multi-modal_2014}) in order to generate the figures.

Figure \ref{fig:veins} displays composite vein images and vein masks for the healthy
volunteer along with the SWI and QSM intermediate derivatives and the
input vein atlas. The derived vein images are comparable to those
generated by the original implementation (\cite{ward_combining_2018}). The complete
workflow took ?? min to complete over ?? compute cores.

@@

Software frameworks (Yacoub and Ammar 2004) have been successful in
improving code quality and efficiency of development in a variety of
contexts (\cite{moore_professional_2008}; \cite{white_hadoop:_2012}; \cite{abadi_tensorflow:_2016}). By
factoring out common elements, only features that are specific to the
given application need to be implemented by the analysis designer, and
the common elements become battle hardened through repeated use. Arcana
handles many of the menial tasks involved with workflow implementation,
such as data retrieval and storage, format conversions, and provenance,
reducing the time and effort required to implement robust workflows from
acquired data to publication results.

An oft-repeated mantra in the open-source software movement dubbed
Linus' Law is that "given a large enough beta-tester and co-developer
base, almost every problem will be characterized quickly and the fix
obvious to someone" or more compactly, "given enough eyeballs, all
\href{https://en.wikipedia.org/wiki/Software_bug}{bugs} are shallow"
(\cite{raymond_cathedral_1999}). Given the size of the neuroimaging research community,
there are a large number of potential beta-testers and co-developers.
However, it has been difficult for researchers to collaborate on the
same code base due to slight differences in acquisition protocols,
storage conventions, researcher preferences, and study requirements.

@@

The abstraction of data and repositories in Arcana enables the same
workflow implementation to be applied to datasets stored in BIDS format
or XNAT repositories. A single code-base can therefore be containerized
into BIDS apps or XNAT pipelines without adaptation, helping to form a
bridge between the two communities of users and developers. Alternative
data storage systems (Book et. al. 2009; Das et. al. 2011, Scott et. al 2011), can be integrated into Arcana by overriding a small number of
methods from the Repository abstract base class. Repository modules
could also be created for data portals such as \emph{DataLad} (Halchenko et. al. 2018) in order to take advantage of the range of platforms they support.
Implementing analyses in Arcana therefore enables researchers and
research groups to easily migrate their workflows between storage
platforms, and not risk being locked in to a particular technology.

@@

Ensuring that the consistent versions of external tools are used
throughout the analysis is important to avoid introducing biases due to
algorithm updates. In systems with environment modules (\cite{furlani_modules:_1991})
installed, Arcana loads and unloads the required modules before and
after each node is executed. When running Arcana within a container,
environment modules and software versions can be installed inside the
container giving exact control over the versions used. To these ends, a
Docker container is available on Docker Hub,
(\href{http://hub.docker.com/r/monashbi/banana}{{http://hub.docker.com/r/monashbi/banana}}),
which can be used as a base for neuroimaging analysis containers.
